Multisequence algorithm for coarse-grained biomolecular simulations: exploring the sequence-structure relationship of proteins

### 2017-10-05 JCP_submit/Manuscript.tex
Final submission and revision

% ****** Start of file aipsamp.tex ******
%
%   This file is part of the AIP files in the AIP distribution for REVTeX 4.
%   Version 4.1 of REVTeX, October 2009
%
%   Copyright (c) 2009 American Institute of Physics.
%
%   See the AIP README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex  aipsamp
%  2)  bibtex aipsamp
%  3)  latex  aipsamp
%  4)  latex  aipsamp
%
% Use this file as a source of example code for your aip document.
% Use the file aiptemplate.tex as a template for your document.
\documentclass[
aip,
%jmp,%
%bmf,%
%sd,%
rsi,%
amsmath,amssymb,
preprint,%
%reprint,%
%author-year,%
%author-numerical,%
]{revtex4-1}

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\title[Multisequence biomolecular simulations]{Multisequence algorithm for coarse-grained biomolecular simulations: exploring the sequence-structure relationship of proteins}

@@

Recent years have seen important advances in biomolecular simulation methods, including improvements to standard molecular dynamics force fields,~\cite{Piana2014} the advent of several alternative atomistic simulation approaches,~\cite{Ding2008,Irback2006,Verma2009,Yang2007} and new techniques for  conformational sampling.~\cite{Bernardi2015} Together with the ever-increasing availability of computational resources, these advances have triggered a few major efforts~\cite{McGuffee2010,Miao2010,Perilla2016,Lindorff-Larsen2011,Yu2016} to characterize the dynamics of biomolecular systems of various sizes, e.g., a small native protein on the millisecond scale~\cite{Lindorff-Larsen2011} and a comprehensive model cytoplasm on the nanosecond scale.~\cite{Yu2016} While encouraging and insightful, these large-scale simulations have also highlighted that severe tradeoffs in size and time scales will likely persist for the foreseeable future.

One way to expand the range of biomolecular simulations is to turn to coarse-grained (CG) models, where the basic aim is to simplify the description of physical interactions while retaining the essential physics of the system under study.~\cite{Riniker2012} Ingolfsson \textit{et al.} list 4 main factors that make CG models computationally fast: reduction in the number of degrees of freedom, faster simulation dynamics, emphasis on short-range interactions and the ability of using larger integration time steps.~\cite{Ingolfsson2014} To this list can be added that a CG representation of either the interaction potential or the molecular geometry often opens up for alternative sampling schemes beyond traditional molecular dynamics approaches, which can further speed up conformational sampling. Examples of such sampling schemes include activation-relaxation kinetics,~\cite{Beland2011} discrete molecular dynamics~\cite{Proctor2011} and various Monte Carlo (MC)-based techniques such as cluster moves.~\cite{Vitalis2009}

@@

As a test application of the MS algorithm, we selected the phenomenon of protein fold switching which recently was demonstrated in a handful of natural and engineered proteins. These special proteins exhibit a unique ability to reversibly switch between entirely different folds, with accompanying changes in secondary structure, hydrophobic core packing and overall shape.~\cite{Bryan2010} The fold switching transitions found in natural proteins typically play a functional role. For example, rare transitions to an alternative fold in the protein KaiB provide a crucial time delay mechanism in the circadian clock cycle of cyanobacteria.~\cite{LiWang2015} Fold switching can occur either spontaneously~\cite{Meier2007} or be triggered by various signals including changes to solution conditions,~\cite{Kuloglu2002} subdomain detachment,~\cite{Burmann2012} ligand binding~\cite{Luo2004} and point mutations.~\cite{Alexander2009} Computational studies, using  coarse-grained~\cite{Holzgrafe2014,Holzgrafe2015,Kouza2012,Sutto2012} or atomistic~\cite{Ramirez-Sarmiento2015,Sikosek2016,Hansen2013} models, have attempted to explain how proteins can exhibit multiple folding funnels and how they are altered in response to binding events or changes in sequence.

The discovery that proteins can be driven to switch folds through an accumulation of point mutations, in particular, holds implications for protein evolution as it suggests a simple mechanism of fold evolution.~\cite{Sikosek2014} Alexander \textit{et al.} demonstrated that the similarly sized but structurally distinct A ($3\alpha$) and B ($4\beta+\alpha$) domains of protein G could, after extensive mutations leaving their respective folds undisturbed, be triggered to switch folds by applying one additional mutation, Y45L, located at the edge of the hydrophobic core in the B domain.~\cite{Alexander2009} This remarkable discovery suggests the possibility that the two domains might be evolutionary related despite a lack of detectable similarity in either sequence or structure in wild-type protein G,~\cite{He2012} although this has yet to be proven. Moreover, it is unclear how common such fold-to-fold transitions are and how they might occur in evolutionary processes.~\cite{Porter2015} In previous work~\cite{Holzgrafe2014,Holzgrafe2015} we showed that mutational paths with abrupt fold switching exist between two other pairs of smaller protein folds within the framework of our CG model.~\cite{Bhattacherjee2012}

In demonstrating mutation-induced fold switching in our model we characterized the folding of a set of 144 different model sequences with 16 amino acids. This set (denoted here {\SR}) was constructed to sparsely span the sequence space between two ideally designed sequences, A1 and N1, folding into an $\alpha$-helix and a $\beta$-hairpin, respectively, as shown in Fig.~2. Here we use the set {\SR} to validate the MS method and compare its computational efficiency to a standard generalized-ensemble method.~\cite{Marinari1992,Lyubartsev1992} We thereafter greatly enlarge {\SR} to a set with 1,024 sequences as well as another set of the same size spanning two 35-amino acid sequences, A2 and TN, that fold into two-helical bundle and mixed $\alpha$-$\beta$ structures, respectively (see Fig.~2). Besides demonstrating that the MS method can be applied to large numbers of sequences, the results allow us to carry out a more systematic analysis of the biophysical properties of sequences along mutational pathways connecting these two pairs of basic folds than has been previously possible.

@@

Conventional Metropolis Monte Carlo simulations of the canonical distribution is problematic at low temperatures for many physical systems because simulations tend to become trapped in local energy minima and hamper representative sampling of configurational space. The basic idea of generalized-ensemble algorithms is to alleviate this trapping problem by sampling states using a non-Boltzmann weight factor and/or expand the state space with additional dynamical parameters~\cite{Mitsutake2001} (for a recent historical account see Ref.~\citenum{Berg2017}). Generalized-ensemble methods that have been frequently used for biomolecular simulations include simulated tempering (ST),~\cite{Marinari1992,Lyubartsev1992} replica exchange~\cite{Swendsen1986}  or parallel tempering,~\cite{Hansmann1997} and metadynamics.~\cite{Laio2002}

ST is a direct extension to the Metropolis algorithm in which the temperature $T$ becomes a dynamic parameter. In this way, frequent visits to high $T$ allow simulations to readily escape from local energy minima. The algorithm thus simulates the joint probability distribution 
\begin{equation}
P(m,\rbar) =\dfrac{1}{\hat{Z}} e^{-\beta_m E(\rbar) + g_m}\,,
\label{ST}
\end{equation}
where  $\beta_m=1/k_\mathrm{B} T_m$, $\{T_m\}_{m=1}^\mathrm{M}$ a set of temperatures and $k_\mathrm{B}$ is Boltzmann's constant. The normalization constant in Eq.~\ref{ST} is  
\begin{equation}
\hat{Z} = \sum_r \sum_{m=1}^{\mathrm{M}}e^{-\beta_m E(\rbar) + g_m}\,,
\end{equation}
where the first sum is over all conformations $\rbar$. The simulation parameters $g_m$ control the marginal probability distribution
\begin{equation}
P(m) = \frac{1}{\hat{Z}}\sum_r e^{-\beta_m E(\rbar) + g_m} \,,
\end{equation}
and must therefore be carefully chosen. A common and convenient choice is $g_m\approx \beta_m F_m$, where $F_m$ is the free energy at temperature $T_m$. With this choice, $P(m)$ becomes approximately flat ensuring all temperatures are frequently visited.

@@

The basic idea of the MS algorithm for biomolecular simulation is to let the sequence $\sbar$ become a dynamic parameter rather than the temperature as in ST. A dynamic $\sbar$ is technically feasible when the potential energy can be written as $E(\sbar,\rbar)$, where $\sbar$ and $\rbar$ are independent variables. This is the case in our coarse-grained protein model which has only backbone degrees of freedom. It can also be achieved in some more detailed models.~\cite{Bhattacherjee2013,Wallin2017}

Similarly to ST, the MS algorithm simulates the joint probability distribution
\begin{equation}
P(\sbar,\rbar) =\dfrac{1}{Z}e^{-\beta E(\sbar,\rbar) + h(\sbar)}\,, 
\label{MS}
\end{equation}
where  
\begin{equation}
Z = \sum_{\sbar\in\mathrm{S}}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)}\,
\label{eq:ZMS}
\end{equation}
and S is a set of pre-selected sequences, i.e., the sequences to which visits are allowed during a simulation. The simulation parameters $h(\sbar)$, similar to the parameters $g_m$ in ST, control the marginal distribution $P(\sbar)=Z^{-1}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)} = \tilde{Z}^{-1}e^{-\beta F(\sbar)+ h(\sbar)}$ and a roughly flat $P(\sbar)$ can be achieved by choosing $h(\sbar) \approx \beta F(\sbar)$, where $F(\sbar)$ is the free energy of sequence $\sbar$ at temperature $T$.

Two types of MC updates are required to sample from the distrubution in Eq.~\ref{MS}, ordinary conformational updates $\rbar\rightarrow\rbar'$ and sequence updates $\sbar\rightarrow\sbar'$. The acceptance probability for the latter  becomes
\begin{equation}
P_\mathrm{acc} (\sbar\rightarrow\sbar') = \min [1, \exp\{-\beta\Delta E+\Delta h\}]\,,
\label{accrej}
\end{equation}
where $\Delta E = E(\sbar',\rbar)-E(\sbar,\rbar)$ and $\Delta h = h({\sbar'})-h(\sbar)$.

{\color{red}Picking a new sequence $\sbar'$ in a sequence update $\sbar\rightarrow\sbar'$ can be done in several ways. One possibility is to draw $\sbar'$ randomly from the set S, such than $\sbar'\ne\sbar$. Alternatively, a type of ``mutational" move can be used where an amino acid position is first picked and then assigned a new amino acid type. The selection of position and type would have to be chosen such that $\sbar'$ does not end up outside S. In this work, we use the former update which is general and guarantees that ergodicity is fulfilled for any S. Importantly, both updates fulfill detailed balance and therefore lead to the same estimates of equilibrium quantities, such as native state stabilities, for the different sequences in S.}

@@

\subsection{Monte Carlo simulation parameters and updates}

Both ST and MS simulations are carried out with two types of conformational updates $\rbar\rightarrow\rbar'$: (1) a global pivot move (20\%) which randomly picks a $\phi_\mathrm{i}$ angle or $\psi_\mathrm{i}$ angle and assign a new value between $-\pi$ and $\pi$; and (2) a semi-local move (80\%) which turns the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$-angles of 4 consecutive amino acids in a coordinated manner.~\cite{Favrin2001}

%, and where S is the set of allowed (pre-selected) sequences, S (see Eq.~\ref{eq:ZMS})
%i.e., given by the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$ angles,

In our MS simulations, sequence updates $\sbar\rightarrow\sbar'$ are carried out in the following way. First, a new sequence $\sbar'$ is picked randomly from the set of pre-selected (allowed) sequences S, such that $\sbar'\ne\sbar$. This new sequence $\sbar'$ therefore differs from $\sbar$ in one or more amino acid positions. Thereafter, the sidechains of the protein, which remains in an unchanged (backbone) conformation $\rbar$, is re-built according to the new sequence $\sbar'$. Practically this means that, at the position(s) where the amino acid type has changed, the sidechain is altered according to the type change. For example, if $\mathrm{p}\rightarrow\mathrm{t}$, the $\mathrm{C}_\beta$ atom is removed and replaced with an $\mathrm{H}_{\alpha 2}$ atom or, if $\mathrm{p}\rightarrow\mathrm{h}$, the $\mathrm{C}_\beta$ remains in place but its character is changed to hydrophobic. Finally, the change in total energy $\Delta E$ is calculated and the accept-reject criterion in Eq.~\ref{accrej} is applied. If rejected, the old state $(\sbar,\rbar)$ is restored.

A sequence update is attempted every $1,000$ MC steps while temperature updates $m\rightarrow m'$ are attempted every 100 steps. The computational cost for sequence updates is somewhat higher than for temperature updates. The latter update does not require any energy calculation and is thus extremely rapid. For the purpose of comparing computational efficiencies of ST and MS, we therefore chose sequence updates to be slightly less frequent than temperature updates while both are fairly frequent. All simulations carried out in this work are summarized in Table~II.

\begin{table}
\caption{\label{tab2} List of simulations carried out in this work. }
\begin{ruledtabular}
\begin{tabular}{lcccr}
Runs & Algorithm & $\kb T$  & MC steps/run\footnote{Excludes a thermalization step with $10^7$ MC steps/run.} &  Sequences\\
\hline
32 & ST & 0.43--0.65 & $1\times10^7$ &A1\\ 
32 & ST & 0.43--0.65 & $1\times10^7$ &N1\\ 
32 & ST & 0.43--0.65 & $1\times10^7$ &R1\\ 
32 & ST & 0.43--0.65 & $1\times10^7$ &R2\\ 
32$\times$8\footnote{32 runs per temperature at 8 different temperatures.} & MS &0.43--0.65& $18\times 10^7$ & $\mathrm{S16}_{144}$\\
16 & MS & 0.43  & $5\times 10^9$ &  $\mathrm{S16}_{1024}$ \\
16 & MS & 0.46 & $4\times 10^9$ &  $\mathrm{S35}_{1024}$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

@@

One might have suspected that the MS algorithm would be hampered by poor acceptance rates for sequence updates. However, this is not the case in our model. We carry out updates $\sbar\rightarrow\sbar'$ by picking a new random sequence $\sbar'\ne\sbar$ from the set of allowed sequences. The (average) acceptance rate  depends on both $T$ and the step in sequences space $\Delta h$, i.e., the number of amino acid positions changed, as shown in Fig.~3. At the lowest $T$ and highest $\Delta h$, acceptance rates are only around 0.1-0.2. However, for most other $T$ and $\Delta h$ the overall acceptance rates are substantially higher and often above the oft-quoted rule-of-thumb value 0.25~\cite{Gilks1996} (see Fig~3B). An increased acceptance rate can easily be achieved by restricting proposed updates such that $\Delta h\le\Delta h_\mathrm{max}$, where $\Delta h_\mathrm{max}$ is a maximum step size, which might be necessary for longer chains. For example, $\Delta h_\mathrm{max} = 1$ would be equivalent to applying only a ``mutational" update, i.e., picking a random (allowed) position and changing the amino acid type at that position.

@@

We now compare the results from our MS calculations with simulated tempering (ST) simulations carried out on 4 of the 144 sequences, namely A1 and N1 and two random sequences, R1 and R2, chosen at distances $h=4$ and $h=6$ from A1, respectively (see Table~I). While ST provides the thermodynamics of a given sequence across a range of $T$ in a single run, an MS simulation provides the thermodynamics of all 144 sequences at one $T$. We adjust the simulation lengths for ST and MS runs such that roughly the same number of sampled conformations are obtained for each $\sbar$ and $T$ combination, thus ensuring that similar computational resources are used for the two algorithms (see Table~II). We first validate the MS algorithm by comparing the average total energy, $\left < E \right >$, calculated for these 4 sequences with the two different methods (see Supplementary Information). The two sets of results are entirely consistent showing that, for a given $\sbar$ and $T$, the MS and ST algorithms indeed sample the same (Boltzmann) distribution.

@@

We have evaluated a biomolecular simulation algorithm that works by making the biological sequence a dynamic parameter. As a test, we applied it on a CG model for protein folding. The results indicate that there are two main benefits of this approach. Firstly, it provides a convenient way to sample the canonical distributions of large numbers of sequences in a single run and, secondly, it enhances the sampling of conformational space meaning it can be applied directly to low temperatures. The conformational sampling efficiency can be assessed from the comparison with ST. Although there is no single ``fair" way to compare the two methods, we chose as a measure of efficiency the statistical error of the total energy, $\sigE$, obtained with roughly the same computational cost per temperature and sequence. At the highest studied temperatures, we find that the statistical errors $\sigE$ are basically the same. This finding is not unexpected because conformational sampling of short polymers at high $T$ does not involve crossing any major energy barriers. As a result, successively sampled conformations for a given combination of sequence $\sbar$ and temperature $T$, are likely uncorrelated in both methods which leads to similar $\sigE$ values.

At lower temperatures, we find that the MS simulations often yields significantly smaller $\sigE$ than ST simulations. It is notable that this acceleration in conformational sampling vis-{\'a}-vis ST is achieved despite that simulations are carried out at constant $T$. Therefore, rather than promoting escape from local minima by visits to higher $T$, as in  ST~\cite{Marinari1992,Lyubartsev1992} or temperature replica-exchange,~\cite{Swendsen1986} MS simulations must escape the minima occurring for one sequence through visits to other sequences. How this process works can be envisioned by considering an MS simulation that is visiting a sequence $s$ and is trapped in a local minimum, requiring that a high free-energy barrier is overcome for escape. The trapped state could be, e.g., a compact $\beta$-sheet rich state with a particular register. Eventually, sequence updates will carry the simulation to other sequences $\hat{s}$ while it conformationally still remains in the trapped state. However, the free energy barrier of escape might be substantially lower for $\hat{s}$ than for $s$, or the barrier may even be altogether absent if the trapped state is unstable for $\hat{s}$, leading to rapid escape from the minimum. The above reasoning also implies that the performance of the MS algorithm likely depends on the size of the sequence set S as well as the conformational properties of the sequences. Specifically, the performance of MS simulations of proteins at low $T$ may benefit from the inclusion of at least a few sequences with poor stability properties, such that partial unfolding of the chain is regularly triggered and thus promoting transitions to new conformational states.

Individual protein sequences that exhibit spontaneous transitions between widely different competing conformational states, such as fold switching proteins, are especially challenging to molecular simulation methods. Representative sampling in such cases requires multiple transitions between highly different states, which can be a slow process. This problem has been addressed by using Hamiltonian replica-exchange techniques to couple G\={o}-like terms, i.e., energy terms with an artificial bias towards a given structure, with a physical force field.~\cite{Meinke2007,Chen2014,Bernhardt2016} This way exchange moves ``feed" diverse conformations into the replica corresponding to the physical force field and enhance sampling.~\cite{Bernhardt2016} We did not specifically study sampling efficiency for sequences exhibiting competing states, such as bistable sequences. It appears likely, however, that a very similar type of ``feeding" of conformations would take place in MS simulations although the coupling occurs instead with other sequences rather than with G\={o}-type terms.

We emphasize that the MS method should not be seen as a general technique to speed up conformational sampling. However, our results indicate that for CG models that permit sequence updates to be carried out as a simple Metropolis step and when visits to higher $T$ is unwanted (or unnecessary), our method can be a highly efficient way to sample the equilibrium behavior of many sequences. This opens up for various applications, such as exploring sequence effects on the conformational properties of disordered~\cite{Das2013} or denatured~\cite{Meng2013} proteins, or as a tool in efforts to combine population genetics and molecular simulations.~\cite{Serohijos2014} While applied to proteins in this work, we note that the theoretical framework of the MS algorithm is equally valid for other bio-macromolecules, including DNA and RNA. The ability to promote conformational sampling without resorting to an increase in $T$ may make the method useful in simulations of biomolecules in ordered phases, such as lipid bilayers or double-stranded DNA, where escape from local minima can be especially challenging~\cite{Bereau2015,Curuksu2009} and elevated $T$ is typically avoided in simulations because it may lead to unwanted perturbations or unraveling of the basic underlying structure.

@@

We have evaluated an algorithm for biomolecular simulations that allows the thermodynamics of multiple  sequences to be calculated in a single run. We applied the algorithm to protein folding and showed that the thermodynamic behavior of $>$1,000 amino acid sequences with up to 35 amino acids could be determined in an intermediate-resolution CG model. The method performs a random walk in sequence space which is especially useful at low temperature as it promotes escape from local minima present in the free energy landscapes of individual sequences. The method might be suitable for CG simulations of various other biomolecular systems, such as peptides in phospholipid bilayers, where sampling at elevated temperatures is not desirable.

\section{Supplementary Material}

See supplementary material for the validation of the MS algorithm.

@@

This work was supported by grants from Memorial University and National Sciences and Engineering Research Council of Canada (RGPIN-2016-05014).

>>>>

% ****** Start of file aipsamp.tex ******
%
%   This file is part of the AIP files in the AIP distribution for REVTeX 4.
%   Version 4.1 of REVTeX, October 2009
%
%   Copyright (c) 2009 American Institute of Physics.
%
%   See the AIP README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex  aipsamp
%  2)  bibtex aipsamp
%  3)  latex  aipsamp
%  4)  latex  aipsamp
%
% Use this file as a source of example code for your aip document.
% Use the file aiptemplate.tex as a template for your document.
\documentclass[
aip,
%jmp,%
%bmf,%
%sd,%
rsi,%
amsmath,amssymb,
%preprint,%
reprint,%
%author-year,%
%author-numerical,%
]{revtex4-1}

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\title[Multisequence Monte Carlo simulations]{Multisequence algorithm for coarse-grained biomolecular simulations: exploring the sequence-structure relationship of proteins}

@@

Recent years have seen important advances in biomolecular simulation methods, including improvements to standard molecular dynamics force fields,~\cite{Piana2014} the advent of several alternative atomistic simulation approaches,~\cite{Ding2008,Irback2006,Verma2009,Yang2007} and new techniques for  conformational sampling.~\cite{Bernardi2015} Together with the ever-increasing availability of computational resources, these advances have triggered a few major efforts~\cite{McGuffee2010,Miao2010,Perilla2016,Lindorff-Larsen2011,Yu2016} to characterize the dynamics of biomolecular systems of various sizes, e.g., a small native protein on the millisecond scale~\cite{Lindorff-Larsen2011} and a comprehensive model cytoplasm on the nanosecond scale.~\cite{Yu2016} While encouraging and insightful, these large-scale simulations have also highlighted the fact that severe tradeoffs in size and time scales will likely persist for the foreseeable future.

One way to expand the range of biomolecular simulations is to turn to coarse-grained (CG) models, where the basic aim is to simplify the physical description of interactions while retaining the essential physics of the system under study.~\cite{Riniker2012} Ingolfsson \textit{et al.} list 4 main factors that make CG models computationally fast: reduction in the number of degrees of freedom, faster simulation dynamics, emphasis on short-range interactions and the ability of using larger integration time steps.~\cite{Ingolfsson2014} To this list can be added that a CG representation of either the interaction potential or the molecular geometry often opens up for alternative sampling schemes beyond traditional molecular dynamics approaches, which can further speed up conformational sampling. Examples of such sampling schemes include activation-relaxation kinetics,~\cite{Beland2011} discrete molecular dynamics~\cite{Proctor2011} and various Monte Carlo (MC)-based techniques such as cluster moves.~\cite{Vitalis2009}

@@

%While this may be hard to achieve for detailed biomolecular models, it is likely possible in many CG models.

As an initial test case, we apply the MS method to a set of 144 model protein sequences with 16 amino acids. This set (denoted here {\SR}) was constructed to sparsely cover the sequence space between two ideally designed sequences, A1 and N1, which fold into an $\alpha$-helix and a $\beta$-hairpin,~\cite{Bhattacherjee2012} respectively, as shown in Fig.~2. In a previous work, this set allowed us to demonstrate the existence of A1-N1 mutational pathways which do not proceed through any disordered (non-folding) sequences but instead switch abruptly between the two folds in a single mutational step.~\cite{Holzgrafe2014} Such an ability to reversibly switch folds was recently demonstrated in a handful of natural and engineered proteins.~\cite{Bryan2010} Here we use the set {\SR} and our CG model for protein folding to validate the MS method and compare its efficiency to a standard generalized-ensemble method.~\cite{Marinari1992,Lyubartsev1992} We thereafter greatly enlarge {\SR} to a set with 1,024 sequences, also spanning the A1-N1 space, as well as another set of the same size spanning two 35-amino acid sequences, A2 and TN, that fold into two-helical bundle and mixed $\alpha$-$\beta$ structures, respectively (see Fig.~2). Besides demonstrating that the MS method can be applied to large numbers of sequences, the results allow us to carry out a more systematic analysis of the biophysical properties of sequences along mutational pathways connecting these two pairs of basic folds than has been previously possible.

@@

Conventional Monte Carlo simulations of the canonical distribution is problematic at low temperatures $T$ for many physical systems because simulations tend to become trapped in local energy minima and hamper representative sampling of configurational space. The basic idea of generalized-ensemble algorithms~\cite{Mitsutake2001} is to alleviate this trapping problem by sampling states using a non-Boltzmann weight factor and/or expand the state space with additional dynamical parameters.

A well-known generalized-ensemble algorithm is simulated tempering (ST),~\cite{Marinari1992,Lyubartsev1992} where it is the temperature that becomes a dynamic parameter. In this scheme, frequent visits to high $T$ allow simulations to readily escape from local energy minima. The ST algorithm simulates the joint probability distribution 
\begin{equation}
P(\rbar,m) =\dfrac{1}{\hat{Z}} e^{-\beta_m E(\rbar) + g_m}\,,
\label{ST}
\end{equation}
where  $\beta_m=1/k_\mathrm{B} T_m$, $\{T_m\}_{m=1}^\mathrm{M}$ a set of temperatures and $k_\mathrm{B}$ is Boltzmann's constant. The normalization constant in Eq.~\ref{ST} is  
\begin{equation}
\hat{Z} = \sum_r \sum_{m=1}^{\mathrm{M}}e^{-\beta_m E(\rbar) + g_m}\,,
\end{equation}
where the first sum is over all conformations $\rbar$. The simulation parameters $g_m$ control the marginal probability distribution
\begin{equation}
P(m) = \frac{1}{\hat{Z}}\sum_r e^{-\beta_m E(\rbar) + g_m} \,,
\end{equation}
and must therefore be carefully chosen. A common and convenient choice is $g_m\approx \beta_m F_m$, where $F_m$ is the free energy at temperature $T_m$. With this choice, $P(m)$ becomes approximately flat ensuring all temperatures are frequently visited.

@@

The basic idea of the MS algorithm for biomolecular simulation is to let the sequence $\sbar$ become a dynamic parameter rather than the temperature as in ST. A dynamic $\sbar$ is technically feasible when the potential energy can be written as $E(\sbar,\rbar)$, where $\sbar$ and $\rbar$ are independent variables. This is the case in our coarse-grained protein model which has only backbone degrees of freedom. It can also be achieved for in some more detailed models.~\cite{Bhattacherjee2013,Wallin2017}

Similarly to ST, the MS algorithm simulates the joint probability distribution
\begin{equation}
P(\sbar,\rbar) =\dfrac{1}{Z}e^{-\beta E(\sbar,\rbar) + h(\sbar)}\,, 
\label{MS}
\end{equation}
where  
\begin{equation}
Z = \sum_{\sbar}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)}\,
\end{equation}
and the first sum goes over a set of allowed sequences $\sbar$. The simulation parameters $h(\sbar)$, similar to the parameters $g_m$ in ST, control the marginal distribution $P(\sbar)=Z^{-1}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)} = \tilde{Z}^{-1}e^{-\beta F(\sbar)+ h(\sbar)}$ and a roughly flat $P(\sbar)$ can be achieved by choosing $h(\sbar) \approx \beta F(\sbar)$, where $F(\sbar)$ is the free energy of sequence $\sbar$ at temperature $T$.

Two types of MC updates are required to sample from the distrubution in Eq.~\ref{MS}, ordinary conformational update $\rbar\rightarrow\rbar'$ and mutational updates $\sbar\rightarrow\sbar'$. The acceptance probabilities for the latter  becomes
\begin{equation}
P_\mathrm{acc} (\sbar\rightarrow\sbar') = \min [1, \exp\{-\beta\Delta E+\Delta h\}]\,,
\label{accrej}
\end{equation}
where $\Delta E = E(\sbar',\rbar)-E(\sbar,\rbar)$ and $\Delta h = h({\sbar'})-h(\sbar)$.

@@

\subsection{Monte Carlo simulation parameters}

Both ST and MS simulations are carried out with two types of conformational updates $\rbar\rightarrow\rbar'$: (1) a global pivot move (20\%) which randomly picks a $\phi_\mathrm{i}$ angle or $\psi_\mathrm{i}$ angle and assign a new value between $-\pi$ and $\pi$; and (2) a semi-local move (80\%) which turns the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$-angles of 4 consecutive amino acids in a coordinated manner.~\cite{Favrin2001} In MS simulations, sequence updates $\sbar\rightarrow\sbar'$ are carried out by randomly picking a new sequence $\sbar'\ne\sbar$ and applying the accept-reject criterion in Eq.~\ref{accrej}. A sequence move is attempted every $1,000$ MC steps while temperature updates $m\rightarrow m'$ are attempted every 100 steps. The simulations carried out in this work are summarized in Table~II.

\begin{table}
\caption{\label{tab2} List of simulations carried out in this work. }
\begin{ruledtabular}
\begin{tabular}{lcccr}
Runs & Algorithm & $\kb T$  & MC steps/run\footnote{Excludes a thermalization step with $10^6$ MC steps/run.} &  Sequences\\
\hline
32 & ST & 0.43--0.65 & $1\times10^7$ &A1\\ 
32 & ST & 0.43--0.65 & $1\times10^7$ &N1\\ 
32 & ST & 0.43--0.65 & $1\times10^7$ &R1\\ 
32 & ST & 0.43--0.65 & $1\times10^7$ &R2\\ 
32$\times$8\footnote{32 runs per temperature at 8 different temperatures.} & MS &0.43--0.65& $18\times 10^7$ & $\mathrm{S16}_{144}$\\
16 & MS & 0.43  & $5\times 10^9$ &  $\mathrm{S16}_{1024}$ \\
16 & MS & 0.46 & $4\times 10^9$ &  $\mathrm{S35}_{1024}$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

@@

One might have suspected that the MS algorithm would be hampered by poor acceptance rates for sequence updates. However, this is not the case in our model. We carry out updates $\sbar\rightarrow\sbar'$ by picking a new random sequence $\sbar'\ne\sbar$ from the set of allowed sequences. The (average) acceptance rate  depends on both $T$ and the step in sequences space $\Delta h$, i.e., the number of amino acid positions changed, as shown in Fig.~3. At the lowest $T$ and highest $\Delta h$, acceptance rates are only around 0.1-0.2. However, for most other $T$ and $\Delta h$ the overall acceptance rate is substantially higher and often above the oft-quoted rule-of-thumb value 0.25~\cite{Gilks1996} (see Fig~4B). An increased acceptance rate can easily be achieved by restricting proposed updates such that $\Delta h\le\Delta h_\mathrm{max}$, where $\Delta h_\mathrm{max}$ is a maximum step size, which might be necessary for longer chains.

@@

We now compare the results from our MS calculations with simulated tempering (ST) simulations carried out on 4 of the 144 sequences, namely A1 and N1 and two random sequences, R1 and R2, chosen at distances $h=4$ and $h=6$ from A1, respectively (see Table~I). While ST provides the thermodynamics of a given sequence across a range of $T$ in a single run, an MS simulation provides the thermodynamics of all 144 sequences at one $T$. We adjust the simulation lengths for ST and MS runs such that roughly the same number of sampled conformations are obtained for each $\sbar$ and $T$ combination, thus ensuring that similar computational resources are used for the two algorithms (see Table~II). We first validate the MS algorithm by comparing the average total energy, $\left < E \right >$, calculated for these 4 sequences with the two different methods (see Supplementary Information). The two sets of results are entirely consistent showing that, for a given $\sbar$ and $T$, the MS and ST algorithms indeed sample the same distribution.

@@

We have evaluated a biomolecular simulation algorithm that works by making the biological sequence a dynamic parameter. As a test, we applied it on a CG model for protein folding. The results indicate that there are two main benefits of this approach. Firstly, it provides a convenient way to sample the canonical distributions of large numbers of sequences in a single run and, secondly, it enhances the sampling of conformational space meaning it can be applied directly at low temperatures. The conformational sampling efficiency can be assessed from the comparison with ST. Although there is no single ``fair" way to compare the two methods, we chose as a measure of efficiency the statistical error of the total energy, $\sigE$, obtained with roughly the same computational cost per temperature and sequence. At the highest studied temperatures, we find that the statistical errors $\sigE$ are basically the same. This finding is not unexpected because conformational sampling of short polymers at high $T$ does not involve crossing any major energy barriers. As a result, successively sampled conformations for a given combination of sequence $\sbar$ and temperature $T$, are likely uncorrelated in both methods which leads to similar $\sigE$ values.

At lower temperatures, we find that the MS simulations often yields significantly smaller $\sigE$ than ST simulations. It is notable that this acceleration in conformational sampling vis-{\'a}-vis ST is achieved despite that simulations are carried out at constant $T$. Hence, rather than promoting escape from local minima by visits to higher $T$, as is the case in  ST~\cite{Marinari1992,Lyubartsev1992} or temperature replica-exchange,~\cite{Swendsen1986}  MS simulations escape local minima through visits to sequences where the minima are not as deep or even altogether absent. For the same reason, we expect the performance of the MS algorithm to depend on the size the sequence set used, as well as their conformational properties.  Specifically, the performance of MS simulations at low $T$ may hinge on the inclusion of at least a few sequences with poor stabilities such that partial unfolding of the chain is regularly triggered and thus promoting transitions to new conformational states.

We emphasize that the MS method should not be seen as a general technique to speed up conformational sampling. However, our results indicate that for CG models that permit sequence updates to be carried out as a simple Metropolis step and when visits to higher $T$ is unwanted (or unnecessary), the MS method can be a highly efficient way to sample the equilibrium behavior of many sequences. In addition, the ability to promote conformational sampling without resorting to an increase in $T$ may make the method useful in simulations of biomolecules in ordered phases, such as lipid bilayers or double-stranded DNA, where escape from local minima can be especially challenging~\cite{Bereau2015,Curuksu2009} and elevated $T$ is typically avoided in simulations because it may lead to unwanted perturbations or unfolding of the basic underlying structure.

@@

We have evaluated an algorithm for biomolecular simulations that allows the thermodynamics of multiple  sequences to be calculated in a single run. We applied the algorithm to protein folding and showed that the thermodynamic behavior of $>$1,000 amino acid sequences with up to 35 amino acids could be determined in an intermediate-resolution CG model. The method performs a random walk in sequence space which is especially useful at low temperature as it promotes escape from local minima present in the free energy landscapes of individual sequences. The method might be suitable for CG simulations of other biomolecular systems, such as peptides in phospholipid bilayers, where sampling at elevated temperatures is not desirable.

@@

This work was supported by grants from Memorial University and National Sciences and Engineering Research Council of Canada (NSERC).
